From noise to knowledge: how randomness generates novel phenomena and reveals information

### 2018-02-22 appendixB/appendixB.Rmd
WIP

---
title: "Appendix B: System size expansion of the master equation for demographic noise"
author: 
  name: "Carl Boettiger"
  affiliation: a
address:
  - code: a
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
layout: 5p  
bibliography: canonical.bib
output: rticles::elsevier_article
preamble: "\\newcommand{\\ud}{\\mathrm{d}}"
---

@@

\begin{align}
\frac{\ud }{\ud t} \langle n \rangle &= \sum n (\mathbb E^{-1} - 1) b_n p_n + \sum n (\mathbb E -1) d_n p_n \nonumber \\ 
&= \sum  b_n p_n(\mathbb E - 1)n  + \sum d_n p_n(\mathbb E^{-1} -1) n \nonumber \\
&= - \langle d_n \rangle +  \langle b_n \rangle \label{mean} \\
\intertext{and similarly the second moment is}
\frac{\ud }{\ud t} \langle n^2 \rangle &= 2 \langle n (b_n - d_n) \rangle + \langle b_n+d_n \rangle \label{var}
\end{align}

The key observation here is that the variance includes the sum of birth and death, not just the difference.  Intuitively we would expect a population with large numbers of births and deaths that average to a small difference to fluctuate much more than one in which this difference is very small.  Despite this simple observation, countless models purporting to include demographic noise ignore this distinction.  While all moments can be derived as above, it is often impossible to solve these equations for nonlinear models.  Instead, we introduce an approximation of Eq \eqref{master1} that is physically motivated and intuitive.  The method is based on derivation presented by the van Kampen [@vanKampen], though a formal proof can be found in earlier work of Thomas Kurtz [@Kurtz1970; @Kurtz1971; @Kurtz1972].

# The system size expansion for a birth-death process

## Introduction, a change of variables

@@

Where we have the part of the birth and death functions that depend on the macroscopic variable $\phi$ alone.  This is known as the macroscopic equation, and corresponds to the density equations commonly written down.  Following @vanKampen, we will call this difference $\alpha_1(\phi)$ as a shorthand\footnote{This notation suggests it is the first jump moment, defined as first moment of the transition rate between states in a Markov process.  It so happens this term will give the macroscopic law for any Markov process, not only a birth-death process.  The second moment of the transition rates, $a_2$, will be for us simply the sum of the birth and death rates.  We use this notation as it generalizes to processes that have a different master equation from \eqref{master1}, to describe a transition of arbitrary step size: $\alpha_{i,j} = \int r^i \Phi_j(r) \ud r$.}

@@

The partial differential equation (PDE) given by \eqref{FP} is sometimes referred to as the Fokker-Planck equation. It is possible to prove this PDE is equivalent to as the stochastic differntial equation (SDE),

which, despite the name, it is important to remember that this is actually an integral equation in the Itô calculus, and not a derivative. Note the separation of the differential elements, which are meant as shorthand for implicit integrals:

The first integral is the ordinary Lebesgue integral, while the second is an Itô integral, which integrates over the stochastic process $B_t$ (Brownian motion, also termed a Weiner process). The Itô calculus follows different rules from the classic calculus, largely due to the fact that an Itô integral has non-zero quadratic variation. (Crudely put, $\textrm{d}x^2 \sim 0$, but $\textrm{d}B^2 \sim dx$).  Though it is common to formulate stochastic models directly as SDEs, it is worth remembering that the SDE is the result of a small noise approximation in the system size expansion, and can break down under larger deviations, even when the mean-field equation remains approximately accurate.  @Ovaskainen2010 provides an overview literature in that addresses this limit and discusses the use of the WKB^[Wentzel, Kramers & Brillouin. Naturally this is KWB in Holland, BWK in France, and JWKB in England (for Jeffreys). Note that the physicist Gregor Wentzel is not the same as the mathematician Alexander Wentzell of the Fredilin-Wentzell theorem for large deviations years later, though I believe the former's result can be derived directly from the latter's.] method to address this case.

@@

Which agrees with calculations elsewhere [@NisbetGurney].

@@

A somewhat different approach to deriving a diffusion equation predominates in the literature.  This approach does not make system size explicit, but instead hinges on taking a simultaneous limit of both short time and small step size.  I prefer the van Kampen derivation, since it appeals to system size rather than an arbitrary small parameter going to zero.  It also provides a more natural treatment for nonlinear systems, for which the diffusion equation requires extra care.  With these notes in mind, the diffusion derivation usually goes as follows:

Consider small steps $x = \epsilon n$, and hence $p_n = \epsilon P_x$, where $\epsilon \ll 1$.  To keep the process from slowing down we consequently have to scale up the rates by $\epsilon$ as well:

For small $\epsilon$ we can ignore the terms of $\mathcal O(\epsilon^2)$, and we have recovered the diffusion approximation.  Unfortunately, nature does not give us a parameter $\epsilon$ that goes to zero.  We expect that larger populations will have relatively smaller fluctuations (recall the introductory example) and hence this limit should have something to do with increasing the system size.  We make these notions more precise by expanding equation \eqref{master1} explicitly in terms of the system size.  This will lead us to a much richer and more satisfying description of demographic noise than the mathematical limit above.

@@

We introduce environmental noise into the master equation by allowing the transition rates to depend explicitly on the macroscopic variable $\psi$ as well as the state variable $\phi$.  It can be helpful to think of fluctuations in $\psi$ as also arising from some lower level transitions, but we need not model these explicitly.  Following the expansion as before (see @vanKampen for details), the mean dynamics are given by the macroscopic equations:

@@

>>>>

---
title: "Appendix B: System size expansion of the master equation for demographic noise"
author: 
  name: "Carl Boettiger"
  affiliation: a
address:
  - code: a
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
layout: 5p  
bibliography: ../paper/refs.bib
output: rticles::elsevier_article
preamble: "\\newcommand{\\ud}{\\mathrm{d}}"
---

@Ovaskainen2010 writes the canonical equation, Eq (1) in the main text, in the form:

$$ \frac{\textrm{d} n}{\textrm{d} t} = \underbrace{f(n)}_{\textrm{det skeleton}} + \underbrace{\sigma_d \xi_d(t) \sqrt{n}}_{\textrm{demographic noise}} + \underbrace{\sigma_e \xi_e(t) n}_{\textrm{environmental noise}}$$

rather than the more formal stochastic differential equation notation shown in the text.  Equations of this form are so-called "Langevin equation," popular in the physics literature [@vanKampen2007].  This notation looks like a more familiar differential equation, but the resemblence is misleading as the standard calculus does not apply to stochastic equations.

A stochastic differntial equation (SDE) of the form

separates out the differential elements$\textrm{d}X_t$, $\textrm{d}t$ and $\textrm{d}B_t$ intentionally to remind the reader that the expression is short-hand not for a derivative expression, but for integrals:

The first integral (over $\textrm{d}t$) is the familiar Reimann integral of the classic calculus, but the second is an Itô integral, which integrates over the stochastic process $B_t$ (Brownian motion, also termed a Weiner process). The Itô calculus follows different rules from the classic calculus, largely due to the fact that an Itô integral has non-zero quadratic variation. (Crudely put, $\textrm{d}x^2 \sim 0$, but $\textrm{d}B^2 \sim dx$).

The intepretation of Langevin equation is less precise, as it can be read to correspond to one of two possible stochastic integrals, and is often associated with a Stratonovich integral.  (Unlike the Itô integral, the Stratonovich integral has zero quadratic variation and so obeys the same chain rule as the classic calculus, but it is not a martingale, e.g. an unbiased random walk, which is a convenient for property for proving theorems).  Fortunately, the notion as a partial differential equation over probabilities does not have this ambiguity, which shows that @Ovaskainen2010 is referring to an Itô expression, just as I have written it more explicitly.  See @vanKampen2007 for an excellent discussion of the Itô-Stratonovich dillema and its intepretation with respect to intrinsic vs environmental noise factors.

## Master equations and the step operator

@@

\begin{align}
\frac{\ud }{\ud t} \langle n \rangle &= \sum n (\mathbb E^{-1} - 1) b_n p_n + \sum n (\mathbb E -1) d_n p_n \nonumber \\ 
&= \sum  b_n p_n(\mathbb E - 1)n  + \sum d_n p_n(\mathbb E^{-1} -1) n \nonumber \\
&= - \langle d_n \rangle +  \langle b_n \rangle \label{mean}
\end{align}

and similarly the second moment is

\begin{align}
\frac{\ud }{\ud t} \langle n^2 \rangle &= 2 \langle n (b_n - d_n) \rangle + \langle b_n+d_n \rangle \label{var}
\end{align}

While all moments can be derived as above, it is often impossible to solve these equations for nonlinear models.  Moreover, we have yet to make any moment closure argument that solving for the first two moments in this manner will be sufficient.  We will illustrate how this is done in both the traditional diffusion approximation and then in the van Kampen system size expansion.

## A traditional derivation

A derivation of the above formulation can be found, e.g. in @Nisbet1982's classic textbook on fluctuating populations, or more recently in @Lande2003.  This diffusion approximation is also given for a more mathematically inclined audience in @Gardiner2009 and in @vanKampen2007.  The approach starts with the same master equation that defines the underlying dynamics of births and deaths (matching the process implemented in the exact Gillespie algorithm, @Gillespie1977, see Appendix A).

The essence of the approach is to argue that in the limit of large populations, the discrete state for the number of indivduals in a population, $N$, can be replaced with a continuous variable, $n$, in which we can expand the master equation as a Taylor series and then truncate at second order.  In contrast to the system size expansion of van Kampen, this approach does not make system size explicit, but instead hinges on taking a simultaneous limit of both short time and small step size.  Consider small steps $x = \epsilon n$, and hence $p_n = \epsilon P_x$, where $\epsilon \ll 1$.  To keep the process from slowing down we consequently have to scale up the rates by $\epsilon$ as well:

For small $\epsilon$ we can ignore the terms of $\mathcal O(\epsilon^2)$, and we have recovered the diffusion approximation.  Unfortunately, we have not been very precise about what we mean by $\epsilon$ -- nature gives us no such obvious small parameter that we can use to "scale down" step sizes and "scale up" rates. The van Kampen expansion makes this notion more precise by introducing the concept of system size.  Instead of counts we consider a change of units into counts per system size, e.g. population density.  We can imagine drawing ever larger rings in our landscape of individuals, defining an ever bigger and bigger "system."

## The System size expansion

 The method is based on derivation presented by the van Kampen [@vanKampen2007], though a formal proof can be found in earlier work of Thomas Kurtz [@Kurtz1970; @Kurtz1971; @Kurtz1972].

### Introduction: a change of variables

@@

Where we have the part of the birth and death functions that depend on the macroscopic variable $\phi$ alone.  This is known as the macroscopic equation, and corresponds to the density equations commonly written down.  Following @vanKampen2007, we will call this difference $\alpha_1(\phi)$ as a shorthand\footnote{This notation suggests it is the first jump moment, defined as first moment of the transition rate between states in a Markov process.  It so happens this term will give the macroscopic law for any Markov process, not only a birth-death process.  The second moment of the transition rates, $a_2$, will be for us simply the sum of the birth and death rates.  We use this notation as it generalizes to processes that have a different master equation from \eqref{master1}, to describe a transition of arbitrary step size: $\alpha_{i,j} = \int r^i \Phi_j(r) \ud r$.}

@@

@@

Which agrees with calculations elsewhere [@Nisbet1982].

@@

@@

We introduce environmental noise into the master equation by allowing the transition rates to depend explicitly on the macroscopic variable $\psi$ as well as the state variable $\phi$.  It can be helpful to think of fluctuations in $\psi$ as also arising from some lower level transitions, but we need not model these explicitly.  Following the expansion as before (see @vanKampen2007 for details), the mean dynamics are given by the macroscopic equations:

@@

## WKB

@Ovaskainen2010 provides an overview literature in that addresses this limit and discusses the use of the WKB^[Wentzel, Kramers & Brillouin. Naturally this is KWB in Holland, BWK in France, and JWKB in England (for Jeffreys). Note that the physicist Gregor Wentzel is not the same as the mathematician Alexander Wentzell of the Fredilin-Wentzell theorem for large deviations years later, though I believe the former's result can be derived directly from the latter's.] method to address this case.

### 2018-02-22 paper/paper.Rmd
WIP

In the extensive literature developing the consequences of both demographic
and environmental noise in models of increasing complexity, it can 
be easy to miss the equally important developments in making the origins
of these sources of noise more precise -- developments which have important
consquences for when and how the resulting models can be applied.

the canonical model on the highly influential @Leigh1981, @Lande1993, @Foley1994

$$ \frac{\textrm{d} n}{\textrm{d} t} = \underbrace{f(n)}_{\textrm{det skeleton}} + \underbrace{\sigma_d \xi_d(t) \sqrt{n}}_{\textrm{demographic noise}} + \underbrace{\sigma_e \xi_e(t) n}_{\textrm{environmental noise}}$$

Stochastic differential equations of this form are so-called "Langevin equation," popular in the physics literature.  This notation is rather sloppy, as the standard calculus does not apply to stochastic equations.  The Langevin formulation, most popular in physics, is usually associated with a Stratonovich integral.  Mathematicians prefer to write stochastic differential equations as $\textrm{d} n = f(x) \textrm{d} t + g(x) \textrm{d} B_t$, where the differentials hint at the underlying definition that this is the sum of a classic Reimann integral (over $\textrm{d}t$) and a stochastic It\^o integral.

Advances in both theoretical understanding and the advent of widely
available and efficient numerical methods have helped drive research
towards increasingly more explicit descriptions of how stochasticity
enters into quantitative models.

@Black2012 provides an important 
review of intrinsic noise and approaches for addressing it, focusing
in particular on use of the van Kampen system size expansion as a
mathematical means to properly bridge between individual-based models
(IBMs, usually implemented only in simulations) and familiar population-level
dynamics (so called macroscopic dynamics).

; @Ovaskainen2010

This division between intrinsic or demographic noise and extrinsic or
environmental noise

The Gillespie simulation provides a simple and exact algorithm
for simulating demographic stochasticity at an individual level.  
Free from both the approximations and mathematical complexity of more common
formulations using stochastic differential equations (SDEs), 
(or their partial differential equation counterparts the Fokker-Planck equation),
the Gillespie algorithm offers a simple and powerful approach to simulating
stochastic phenomena that remains underutilized.

The mathematical proofs @Kurtz1970 & @Kurtz1971, popularized by
@vanKampen's intuitive "system size" expansion, provide a far more 
explicit framework to interpret demographic stochasticity than the
better known diffusion approximation [@Nisbet1982 @Lande2003; 
see @Gardiner2009].

@@

@@

It is possible to simulate such a process exactly using Gillespie's
exact stochastic simulation algorithm [SSA; @Gillespie1977], which
simulates each individual birth or death as a separate event.

@@

Figure \ref{gillespie}
uses this algorithm to compare simulations of @Levins1969 patch model;
(see Appendix A for example code).  In the Levins'  model,

@@

The van Kampen system size expansion provides a convenient way to
explicitly approximate the "macroscopic" (i.e. population level,
see @Black2012) dynamics from an individual-based description of
events in a Markov process [@vanKampen2007; based on @Kurtz1970;
@Kurtz1971].  In addition to illustrating how common stochastic
models (in particular, Gaussian noise) arise from lower-level
descriptions, this result will also provide us a useful framework
for interpreting the behavior of stochastic models more generally.
In Appendix B I illustrate the derivation of the following equations
for the mean and expected population sizes for a generic density
dependent birth-death process:

@@

>>>>

@@

In the extensive literature developing the consequences of both demographic
and environmental noise in models of increasing complexity, it can 
be easy to miss the equally important developments in making the origins
of these sources of noise more precise. Models in which stochasticity was
essentially "tacked on" to a deterministic skeleton can be substituted for 
approaches that permit both the skeleton and noise to emerge from underlying
processes.  This more mechanistically specific approach to the origins of 
noise can indeed recover the classic formulations, but it can also 
help highlight where those models break down.
Much of the literature [e.g. see @Ovaskainen2010] has focused on the following
model, established in highly influential papers of @Leigh1981, @Lande1993,
@Foley1994:

$$ \textrm{d} N_t = \underbrace{f(N_t) \textrm{d} t}_{\textrm{det skeleton}} + \underbrace{\sigma_d \sqrt{N_t} \phantom{\cdot}\textrm{d}B_t^{(d)} }_{\textrm{demographic noise}} + \underbrace{\sigma_e N_t \phantom{\cdot}\textrm{d}B_t^{(e)}}_{\textrm{environmental noise}}$$

Where $\textrm{d}B_t^{(d)}$ and $\textrm{d}B_t^{(e)}$ refer to Brownian processes, (i.e. Gaussian white noise process) for demographic and environmental stochasticity respectively.  When $f(n)$ is simply a Verhulst logistic function this is often referred to as the canonical model [e.g. @@Ovaskainen2010], and provides a clear illustration of the partition of separate deterministic and stochastic elements. Significant research continues to focus on extending the study of this model to consider more complex and higher-dimensional $f(n)$ (coexistence models, stage structured models) and consider the case of autocorrelated, rather than than white noise of $\textrm{d}B_t^{(e)}$.

Before we follow that path, I will review how a shift towards more mechanistic descriptions of the origins of noise provide a more nuianced view of when the canonical model does an does not hold in the context of both demographic and environmental stochasticity.  My approach will highlight two fundamental tools: an ability for efficient, exact simulation using @Gillespie1977's stochastic simulation algorithm (SSA), and the van Kampen system size expansion [@vanKampen2007], which is rooted in the mathematical theorems of @Kurtz1970 & @Kurtz1971, in place of the more popular diffusion approximation derived by Taylor series [e.g. @Nisbet1982 @Lande2003; see comparisons in @vanKampen2007 and @Gardiner2009].

@@

The @Gillespie1977 provides a simple and exact algorithm
for simulating demographic stochasticity at an individual level.  
Free from both the approximations and mathematical complexity,
the Gillespie algorithm offers a simple and powerful approach to simulating
stochastic phenomena that remains underutilized.

@@

@@

The van Kampen system size expansion, shown for a general birth-death
process in Appendix B, derives an ordinary differential equation (ODE)
for the macroscopic process and a *linear* Fokker-Planck equation which
governs the *deviations* from the macroscopic equation:

@@

Note that the van Kampen expansion is not identical to the SDE 
postulated by the canonical equation, which corresponds to a 
non-linear Fokker-Planck equation for the probability density.  
As shown in Appendix B, the popular expansion
